\bf Model Visualization Techniques for a Social Network Model

### 2017-03-13 paper.Rnw
beefing up section 3

\section{Introduction}

@@

\st{
\begin{itemize}
\item What? - model visualization and SAOMs. briefly dsecribe what these things are
\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
\item How?
\end{itemize}
also talk about other dynamic network models briefly to put SAOMs in context.}

The guiding principles of model visualization from \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result. Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization. First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process. Bringing the CTMC out from behind-the-scenes and into the light of visualization  provides researchers with insights into the underlying features of the model. Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which are attached to a network statistic. These statistics are often somewhat, if not highly, correlated, and so are the associated parameters. By visualizing collections of SAOMs, we can gain a better understanding of these correlations and find ways to deal with them and rectify their effects. Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks rely on simulations from the fitted model. Again, in these steps are largely hidden from the researcher computing them with software. We hope that by bringing these background elements and others into the foreground, stochastic actor-oriented models will be better understood and more accessible to social network analysts.

@@

When node $i$ is given the chance to change a node, we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from "the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p. 368]{snijders01}. This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of

@@

@@

@@

\st{
\begin{itemize}
\item WHat is the model? the network model (SAOM model is just one of many)
\item What is the data? Edge data, node data, observed time points, unobserved timepoints (CTMC)
\item Note: specifically dealing with dynamic networks observed at discrete time points. 
\end{itemize}
What are some things that could go in here? need more data examples; not just the friends data.
Harry Potter data??? I want to find my own data, too.}

@@

The second principle of Wickham et al states that ``collections are more informative than singletons" (p. 210). They describe several different methods for producing such collections, but we focus on only a few here. We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model." (p. 210-11) We also fit the same model many times, resulting in distributions of fitted parameters for one data set.

@@

library(geomnet)
M1avg$from <- paste0("n", M1avg$from)
M1avg$to <- paste0("n", M1avg$to)
M1avg2$from <- paste0("n", M1avg2$from)
M1avg2$to <- paste0("n", M1avg2$to)

>>>>

\section{Introduction} 
\st{
\begin{itemize}
\item What? - model visualization and SAOMs. briefly dsecribe what these things are
\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
\item How?
\end{itemize}
also talk about other dynamic network models briefly to put SAOMs in context.}

@@

\par The software SIENA, and it's \texttt{R} implementation \texttt{RSiena}, was developed by \citet{RSiena} to estimate the parameters of the SAOMs, among other social network models. This software is a huge contribution to the field of social network analysis, and when we began working with it we were eager to understand how it works. To better understand the \texttt{RSiena} software and the SAOMs themselves, we use the principals of model visualization from \citet{modelvis}. With the work of \citet{modelvis} guiding us, we explore RSiena and SAOMs with innovative visualizations, both static and dynamic. By bringing some light to the underlying methods and structures "behind the scenes" of SAOMs and RSiena, we will help other researchers working with the models better understand and better use these models.

\par The guiding principles of model visualization from \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result. Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization. First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process. Bringing the CTMC out from behind-the-scenes and into the light of visualization  provides researchers with insights into the underlying features of the model. Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which are attached to a network statistic. These statistics are often somewhat, if not highly, correlated, and so are the associated parameters. By visualizing collections of SAOMs, we can gain a better understanding of these correlations and find ways to deal with them and rectify their effects. Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks rely on simulations from the fitted model. Again, in these steps are largely hidden from the researcher computing them with software. We hope that by bringing these background elements and others into the foreground, stochastic actor-oriented models will be better understood and more accessible to social network analysts.

@@

When node $i$ is given the chance to change a tie, we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from "the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p. 368]{snijders01}. This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of

@@

To fit a SAOM to longitudinal network data, we use the package \texttt{RSiena} \citep{RSiena}. This package uses simulation methods to estimate parameter values using either the method of moments or maximum likelihood estimation. In this paper, we focus of the method of moments estimation because the theory behind it was established in \citep{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}.

@@

The concept of model visualization was developed to complement traditional model diagnostic tools. Typically, numerical summaries such as $R^2$ are used to assess model fit, and visualizations, like residual plots, are restricted to the data. Every good data analysis includes both numerical and visual summaries of the data, so why should model description be restricted to numerical summaries? \citeauthor{modelvis} outline the three different parts of a model: the model family, the model form, and the fitted model. The latter is primarily what one thinks of first when considering a model in a data analysis, where a model is fit to the data, and parameter estimates reported. In the context of SAOMs, the fitted model contains the estimated rate parameters and the estimated objective function parameters. The model form describes the the model before the fitting process and how variables describe the data in the model. In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. Finally, the model family is the most broad description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem and knowledge at had. For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network dynamics. These three parts of a "model" are should be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.

@@

In \citeauthor{modelvis}, they define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. What does this definition mean for network data? For dynamic social networks, there are a few different data ``spaces." First, we have the actors and their corresponding covariates. Second, we have the dyad variables that describe the relationships between the variables. Finally, we have the time over which the network is allowed to evolve. These three data pieces can be visualized together in various ways. The traditional network visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if the dyad between two nodes has value greater than zero, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side. 
\st{
\begin{itemize}
\item WHat is the model? the network model (SAOM model is just one of many)
\item What is the data? Edge data, node data, observed time points, unobserved timepoints (CTMC)
\item Note: specifically dealing with dynamic networks observed at discrete time points. 
\end{itemize}
What are some things that could go in here? need more data examples; not just the friends data.
Harry Potter data??? I want to find my own data, too.}

@@

The second principle of Wickham et al states that ``collections are more informative than singletons" \citep[p.~210]{modelvis}. They describe several different methods for producing such collections, but we focus on only a few here. We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}. We also fit the same model many times, resulting in distributions of fitted parameters for one data set.

@@

library(geomnet)
M1avg$from <- paste0("V", M1avg$from)
M1avg$to <- paste0("V", M1avg$to)
M1avg2$from <- paste0("V", M1avg2$from)
M1avg2$to <- paste0("V", M1avg2$to)
